Combining Language Resources Into A Grammar-Driven Swedish parser

### 2012-03-05 doc/paper/PaperFin.tex
second draft of final lrec paper

\address{ Department of Computer Science \& Engineering, Gothenburg University \\
               Address1 \\
               author1@xxx.yy, author2@zzz.edu, author3@hhh.com\\}

% 150-200 words here
\abstract{This paper describes on-going work on a rule-based, open-source parser for
Swedish. The central component is a wide-coverage grammar implemented in the GF
formalism (Grammatical Framework), a dependently typed grammar formalism
based on Martin-L{\"o}f type theory. GF has strong support for multilinguality and
has so far been used successfully for
controlled languages \cite{cnl} and recent experiments have showed
that it is also possible to use the framework for parsing free language \cite{patent}.
In addition
to GF, we use two other main resources: the Swedish treebank Talbanken and the
electronic SALDO lexicon. By combining the grammar together with
a lexicon extracted from SALDO we obtain a parser accepting all sentences
described by the given rules. We develop and test this
on examples from Talbanken.
The resulting parser gives a full syntactic analysis of the input sentences.
It will be highly reusable, freely available, and as GF provides libraries for
compiling grammars to a number of programming languages, chosen parts of the the 
grammar may be used in various NLP applications. 
\\ \newline \Keywords{GF, Swedish, Parsing}}

@@

\section{Introduction}
% Copied from Master's Thesis. ok to do that?
Swedish is a North-Germanic language,
closely related to Norwegian and Danish. The languages share most of their
grammatical structures and are mutually intelligible. Swedish is also 
one of the official languages in Finland and altogether spoken by approximately 9
million people.
Swedish syntax is often similar to English, but the  morphology is richer and the
word order slightly more intricate.
It is a verb-second language: the second constituent of a declarative main
clause must consist of a verb.
The first constituent of the clause is usually made up of the subject,
although it likewise could consist of adverbial phrases or objects.
Fronting the finite verb marks questions.

Our goal is to implement a wide-coverage grammar and parser for Swedish
using the GF formalism, and 
thereby investigate how GF can be used for open-domain parsing.
We compile a large-scale grammar to a parser
and combine it with the extensive lexicon SALDO. The Swedish treebank 
Talbanken provides manually tagged trees which we use for improving and evaluating
the grammar. The parser will additionally
be evaluated by experts. \\
We chose our resources so that both our grammar and parser can 
be freely available and open-source.

This paper will briefly introduce GF, Talbanken and SALDO in Section \ref{sec:background}
and describe our results in Section \ref{sec:progress}.

@@

Grammatical Framework \cite{ranta-2011} (GF) is a grammar formalism based on
functional programming.

@@

Since many of the languages in the GF library resemble each other grammatically,
they can share much of their implementations. This is usually done by using a
\verb|Functor|, avoids code duplication and aids the code maintenance.

GF has so far been used in a number of projects,
MOLTO\footnote{http://www.molto-project.eu/}, TALK \cite{talk}
and WebAlt \cite{webalt} to mention a few. 
All those are special domain applications, dealing with controlled natural
language.
This project takes a different approach by using GF for open domain language.
Using the Swedish resource grammar as our starting
point, we get a basic description of the language. The framework provides
tools such as parsing, generation and
a well-tested interpretation of the parse trees. Furthermore, there are tools
for using GF grammars in a number of programming languages like Haskell
and Java.

\subsection{Talbanken}
For development and evaluation, we use the Swedish treebank
Talbanken \cite{talbanken}.
It was assembled in the 1970s at Lund University and modernized
in 2005 \cite{talbanken05} and
enriched with annotation for a full phrase structure analysis.  \\
Although Talbanken contains both written and spoken Swedish,
only the prose material, consisting of 6316 sentences, will be used in this project.
This part was also used when training the data-driven parser Maltparser \cite{malt}. \\

\subsection{Saldo}
SALDO \cite{saldo} is an open source lexicon resource
based on Svenskt Associationslexikon. It is
developed at Spr\{aa}kbanken at Gothenburg University
and intended for usage in language technology
research. 
We are developing tools for extracting GF lexicons
from SALDO, which have resulted in a dictionary containing
over 98000 entries including information about verbs that
are reflexive or needs a particle.
The importing method should be fast and reliable
enough to allow us to always have a fresh version of the dictionary
in GF.

\section{Related work}
%Copied form thesis!!!! 
% Add: how are these related to this work?
% Add :detaild comparison of outputs of various parsers of Sw.
Many years of research have lead to many interesting language
technology tools for Swedish.
An example is the well-known data-driven Malt parser \cite{malt},
which has been trained on Talbanken. 
There are also a number of grammar-based parsers, although none is freely available.
The cascaded finite state parser CassSwe \cite{casswe} and
The Swedish Constraint Grammar \cite{birn}
give syntactic analyses. 
Swedish FDG (Voultanien,2001) uses the Functional Dependency Grammar
\cite{fdg},
an extension of the Constraint Grammar
formalism, and produces a dependency structure focusing on finding the nominal
arguments. \\
The LinGO Grammar Matrix \cite{matrix}, is a starter-kit for building
Head-Driven Phrase Structure Grammars \cite{hpsg} (HPSG) providing
compatibility with tools for parsing, evaluation, semantic representations etc.
Translation is supported by using Minimal Recursion
Semantics \cite{mrs} as an interlingua. \\
There is a collection of grammars implemented in this framework, giving broad-coverage
descriptions of 
English, Japanese and German. 
The Scandinavian Grammar Matrix \cite{scandmatrix} covers common parts of
Scandinavian, while Norsource (Hellan, 2003) describes Norwegian. A Swedish version
was based upon this (SweCore, Ahrenberg) covering the morphology and some
differences between Swedish and Norwegian. Further, there is the BiTSE 
grammar \cite{stymne}, also implemented using the Lingo Matrix,
which focuses on describing and translating verb frames.\\

@@

\subsection{Extracting a large lexicon}
The lexicon provided with the GF resources is far too small for open-domain
parsing.
This section describes the process of importing Saldo, which is 
compatible with GF, and easily translated to GF format.
As Saldo is continuously updated, the importing process has been designed to be fast
and stable enough to be redone at any time.

\subsubsection{Implementation}
The basic algorithm for importing Saldo was implemented by Angelov (2008)
and  produces code for a GF lexicon.
For each word in Saldo, it decides which forms should be used as input
to the GF smart paradigms. For a verb, this will in most cases mean giving
the present tense form, see figure \ref{fig:saldoknyt}. \\

@@

All assumed paradigms are printed to a temporary lexicon, 
which will produce an inflection table for every entry when compiled.
The tables are compared to the information given
in Saldo and if the tables are equal the code for the word is saved. If the table
is erroneous, another try is made
by giving more forms to the smart paradigm.
For example \ref{fig:saldoknyt}, the smart paradigm will fail to calculate the
correct inflection table. In the next try both the present and the past tense
are given:\\

\begin{figure}[h]
\begin{center}
\verb-mkV "knyter" "kn{\"o}t" ;-
\caption{Second output for the verb `knyta'}
\label{fig:saldoknyt2}
\end{center}
\end{figure}

The program is run iteratively until the GF table matches the one given in Saldo,
or until there are no more ways of using the smart paradigm. The verb 'knyta'
will need three forms:\\

\begin{figure}[h]
\begin{center}
\verb-mkV "knyter" "kn{\"o}t" "knutit"-\\
\caption{Final output for the verb `knyta'}
\label{fig:saldoknyt3}
\end{center}
\end{figure}

\subsubsection{Results}
\label{sec:saldoRes}
The resulting dictionary contains more than 100 000 entries, approximately 80 \% 
of the total size of Saldo.
There are a number of reasons why some words were not imported,
the most obvious one is that we do not want all categories from
Saldo in the GF lexicon. Prepositions, 
numerals, 
personal pronouns etc.
are assumed to be present in the resource grammars and should not be added again.
Saldo contains many pronouns which
are not analysed the same way in GF %TODO see where??
(see section \ref{sec:gfnouns}).
Before adding them to our lexicon, we need to do more analysing to find their
correct GF-category. 
%Some experiments on finding the category
%have been done using Talbanken, see section \ref{sec:gf.quant}.

Categories involving multiple words 
are usually handled as idioms and should be given in a separate lexicon. In
total six types of words were considered for the extraction: \\

\begin{figure}[h]
\begin{tabular}{|l|ll|}
\hline
& Saldo$/$ GF tag & Example \\
\hline
 Adverb & \textbf{ab} $/$\textbf{Adv} & ofta (\emph{often})\\
 Adjective&\textbf{av} $/$    \textbf{A} & gul (\emph{yellow})\\
 Noun & \textbf{nn} $/$\textbf{N} & hus (\emph{house})\\
 Verb & \textbf{vb} $/$\textbf{V} & springa (\emph{run})\\
 Reflexive verbs  & \textbf{vbm}$/$\textbf{V} & raka sig (\emph{shave})\\
 Verbs with particles &\textbf{vbm}$/$\textbf{V}  &  piggna till (\emph{perk up})\\
\hline
\end{tabular}
\caption{}
\end{figure}

Most but not all words of these categories have been imported.
One reason why the importing phase would fail 
is that Saldo, unlike GF, only contains the actually used word forms.
For technical reasons, the smart paradigm might need forms never used.
Consider for example the plural tantum 
noun \emph{`glas{\"o}gon'} (\emph{`glasses'}).
The smart paradigm requires a singular form, and since the program could not
find this in Saldo, there was no way of adding the lemma to the lexicon. 
When the program failed to import a noun, this was often the explanation.
Words of this type may be added manually, for \emph{`glas{\"o}gon'} we could use
the ostensibly correct singular form `glas{\"o}ga', although this
has another meaning (`glass-eye').
The same problem occurred for the irregular s-verbs,
(\emph{`synas'} \emph{(`show')} or \emph{umg{\aa}s} \emph{(`socialize')})
which made up 61.5 \% of the failing verbs of type \verb_vb_.\\
In a few cases the smart paradigms could not generate the correct declination.\\

When testing the coverage of Talbanken,
we found that there are around 2500 word forms still missing, excluding the ones
tagged as names and numbers. This number may seem very high, but 
4/5 of the word forms are compounds and when performing the intended parsing,
an additional analysis identifying compounds should be preformed before
looking-up the words in the lexicon. We
should also take into consideration that 
we cannot automatically find out how many actually stem from the same word, or
how many abbreviations that are present. Talbanken also contains a small number of 
spelling errors, which probably are enumerated among our missing words. The majority
of the missing words are only used once.\\

@@

A list of words that were given different labels in GF than in Talbanken has been
composed, consisting of about 1600 entries. Many of those are
acceptable and reflects the difference 
made in the analyses, such as the examples in figure \ref{tab:saldodiff}.
Others are examples of words that are still missing from
the lexicon.

%Since Saldo does not give any valency information, neither does the imported lexicon.
%The lexicon does not contain any information.
\noindent Valency information, which is crucial for GF, is not given in Saldo and
hence not in the imported lexicon. Instead we extract this from
Lexin\footnote{http://spraakbanken.gu.se/lexin/}.

\subsection{A tool for lexical acquisition}
As we extract our main lexicon from Saldo, we have also created a tool for
semi-automatic acquisition to complement the lexicon. We have
tested it on verbs, with good results. It makes use of
the \emph{smart paradigm} given in the Swedish resource grammar.
The smart paradigm is a function which given one form of a word, can
infer which paradigm it most likely belongs to.
For verbs, the paradigm accepts words in present tense indicative form.
If needed, it also accepts more verb forms showing the correct inflection.

By combining this method with the information from the tags in Talbanken,
the tool interactively generates GF lexical entries. 
Given a list of words, it iteratively
tries to figure out how to conjugate each of them. If several forms of a word are 
given, the program will try to identify the one that carries the most linguistic
information, put this in a form recognized by the smart paradigm and ask GF to output
a table with the resulting inflection. 
If the table contains all other conjugations from the input list,
the program will ask the  user to
validate the claimed paradigm. The user may now either
allow the word to be added to the lexicon, remove it or request another guess.
The user hence only needs to decide if each paradigm is correct or not, and
the program will do the rest.

The tool has been tested on verbs. Although using simple techniques it 
manages to assign the correct paradigm to 70-75 \% of the given lemmas.
A smaller test has shown that out of the accepted lemmas, the correct guess is
made directly in 75\% of the cases, whereas the user has to reject one or more
guesses for 25\%. Hence the program gets a correctness rate of 56 \% if it is
run without interference from the user. This kind of use is however not intended.

@@

\section{Translating The Treebank Talbanken}
The information from the tags in Talbanken can be used for many purposes.
We are currently working on an automatic transformation of Talbanken trees 
to trees in GF format. The translation makes use of the POS tags as well as
the syntactic information, and the mapping has so far turned out to be unambiguous.

The mapping gives information about which form a word is
currently used in and this may be used by the lexical extraction
tools.
The translated trees enable us to extract probabilities for how often
different functions are used, a feature that will enable disambiguation.
Furthermore, the translation makes it easy to identify grammatical constructions
missing from the GF grammar and shows how the GF analysis differs from the one made
in Talbanken.
Another important use of the mapping is evaluation of the parser, which can be
accomplished by comparing the parse trees and the trees from the transformer.

We have so far concentrated on shorter sentences, without idioms and
conjunction. 
Since GF demands information 
that is not given by the Talbanken tags -- such as the valency of verbs -- a
full translation requires that all words used are already known to GF. 
The evaluation of trees of this sort shows that the mapping
can restore 85\% of the nodes. 
If we lift the restrictions, we can restore 63\% of the tags.

\section{Development of the grammar}
An important part of this project has been to develop the Swedish GF grammar and
to adapt it to cover constructions used in Talbanken. As a grammar
implementation can never be expected to give
full coverage of a language, we aim for a grammar fragment which gives a deep
analysis of the most important Swedish constructions.
The starting point has been the GF resource grammar and the new implementation is
still compatible with this.
It has earlier been hard to identify missing constructions of the Swedish
implementation, since there was no large resource available to evaluate it on.

Our evaluations are based on Talbanken, and when first conducting tests,
we found much room for improvement.

For Swedish, about 85\% of the GF resource code is shared with the other Scandinavian
languages. 
However, if we aim for a deeper and more comprehensive analysis of Swedish,
the implementation of the languages needs to be more independent.
The resource grammar gives a good start and our present grammar 
% don't say inverted word order
covers constructions such as declarative sentences with normal or inverted
word order, questions, passives, imperatives, relative clauses, cleft
constructions etc.
A number of constructs that are generally not present in other languages and
therefore not given by the resources, have also been added. 
%These include the use of
%the reflexive pronoun \emph{sitt}: \\
%\emph{Han s{\aa}g \textbf{sitt} hus} $\; \; \; \;$ (\emph{He saw his (own) house}) $\;$
%as opposed to \\
%\emph{Han s{\aa}g \textbf{hans} hus} $\;$ (\emph{He saw his (another person's) house}) \\
%Fronting words or phrases is very common in Swedish and are now allowed by the grammar:\\
%\emph{\textbf{Glad} var han inte.} $\;$ (\emph{\textbf{Happy} was he not}). \\
%This sort of rephrasing is not given by the resource grammar, since it
%has little effect on the logical representation.
\subsection{The s-passive}
Passive voice is often used in Swedish, especially the
\textit{s-passive}.
\enumsentence{
\shortex{5}{Uppsatsen & skrev\textbf{s} & av & en & student.}
        {\emph{the essay} & \emph{wrote+\textbf{s}} & \emph{by} & \emph{a} & \emph{student}}
        {``The essay was written by a student."}}\label{sent:skrevs}
%{b. &En & student & skrev & uppsatsen.}
%        {&a &student &wrote &the essay}
%        {~~`A student wrote the essay.'}}
Some studies suggest that the s-passive is used in more than 80 \% of the times
\cite{laanemets}.
It is however not as common in the other Scandinavian languages, 
The resource grammar for Scandinavian therefore implemented a function for passive,
\verb-PassV2-, by using auxiliary verb. \\

During this project, the s-passive was added although the
periphrastic passive is still allowed.
The grammar further allows not only two-place verbs, but all verb phrases missing an object, to 
form passive:
\enumsentence{\textbf{Active use of two-place verb}\\\shortex{4}
{Vi & erbj{\"o}d& henne& jobbet }
{\emph{we} & \emph{offered}& \emph{her}&\emph{the job}}
{``We offered her the job"}}
\label{sent:give2pass}
\enumsentence{\textbf{First place in two-place verbs}\\\shortex{3}
{Hon & erbj{\"o}ds & jobbet}
{\emph{she} &\emph{ offered+\textbf{s}}&\emph{ the job}}
{``She was offered the job"}}
\label{ex:passV33}
\enumsentence{\textbf{Second place in two-place verbs}\\\shortex{3}
{Jobbet & erbj{\"o}ds & henne }
{\emph{the job} & \emph{offered+\textbf{s}}&\emph{ her}}
{``The job was offered to her"}}
\label{ex:passV32}
\enumsentence{\textbf{V2A verb}\\\shortex{3}
{Huset & m{\aa}lades & r{\"o}tt}
{\emph{the house} & \emph{painted+\textbf{s}}& \emph{red}}
{``The house was painted red"}}
\label{ex:passV2A}

>>>>

\address{ Department of Computer Science \& Engineering\\
          Gothenburg University \\
          Box 8718, SE-402 75 Gothenburg, Sweden\\
          ahlberg.malin@gmail.com, enache@chalmers.se \\}

% 150-200 words here
\abstract{This paper describes work on a rule-based, open-source parser for
Swedish. The central component is a wide-coverage grammar implemented in the GF
formalism (Grammatical Framework), a dependently typed grammar formalism
based on Martin-L{\"o}f type theory. GF has strong support for multilinguality and
has so far been used successfully for
controlled languages \cite{cnl} and recent experiments have showed
that it is also possible to use the framework for parsing free language.
In addition
to GF, we use two other main resources: the Swedish treebank Talbanken and the
electronic SALDO lexicon. By combining the grammar with
a lexicon extracted from SALDO we obtain a parser accepting all sentences
described by the given rules. We develop and test this
on examples from Talbanken.
The resulting parser gives a full syntactic analysis of the input sentences.
It will be highly reusable, freely available, and as GF provides libraries for
compiling grammars to a number of programming languages, chosen parts of the the 
grammar may be used in various NLP applications. 
\\ \newline \Keywords{GF, Swedish, Parsing}}

@@

\section{Introduction}
% Copied from Master's Thesis. ok to do that?
Our goal is to implement a wide-coverage grammar and parser for Swedish
using the GF formalism, and 
thereby investigate how GF can be used for open-domain parsing.
We compile a large-scale grammar to a parser
and combine it with the extensive lexicon SALDO. The Swedish treebank 
Talbanken provides manually tagged trees which we use for improving and evaluating
the grammar. The parser will additionally
be evaluated by experts. \\
We chose our resources so that both our grammar and parser can 
be freely available and open-source. 
The target language is Swedish, a North-Germanic language
closely related to Norwegian and Danish. The languages share most of their
grammatical structures and are mutually intelligible. Swedish is also 
one of the official languages in Finland and altogether spoken by approximately 9
million people.
Swedish syntax is often similar to English, but the  morphology is richer and the
word order slightly more intricate.
It is a verb-second language: the second constituent of a declarative main
clause must consist of a verb.
The first constituent of the clause is usually made up of the subject,
although it likewise could consist of adverbial phrases or objects.
Fronting the finite verb marks questions.

This paper will briefly introduce GF, Talbanken and SALDO in Section \ref{sec:background}
Section \ref{sec:extractsaldo} explains how we extract a lexicon, section \ref{sec:mapping}
how we translate Talbanken to GF annotation and section \ref{sec:grammar}
presents the work on the grammar implementation.

@@

Grammatical Framework \cite{ranta-2011} is a grammar formalism based on
functional programming.

@@

Since many of the languages in the GF library resemble each other grammatically,
they can share much of their implementations. This is usually done by using a
\verb|Functor| by which we avoid code duplication and which aids the code maintenance.

GF has so far been used in a number of projects,
MOLTO\footnote{http://www.molto-project.eu/}, TALK \cite{talk}
and WebAlt \cite{webalt} to mention a few. 
All those are special domain applications, dealing with controlled natural
language.
This project takes a different approach by using GF for open domain language,
as done in recently conducted work on translating patents \cite{patent}.

Using the Swedish resource grammar as our starting
point, we get a basic description of the language. The framework provides
tools such as parsing, generation and
a well-tested interpretation of the parse trees. Furthermore, there are tools
for using GF grammars in a number of programming languages like Haskell
and Java.

\subsection{Talbanken}
For development and evaluation, we use the Swedish treebank
Talbanken \cite{talbanken}.
It was assembled in the 1970s at Lund University and modernized
in 2005 \cite{talbanken05} and
enriched with annotation for a full phrase structure analysis.  \\
Although Talbanken contains both written and spoken Swedish,
only the prose material, consisting of 6316 sentences, is used in this
project.
This part was also used when training the data-driven parser Maltparser \cite{malt}. \\

\subsection{SALDO}
SALDO \cite{saldo} is an open source lexicon resource
based on Svenskt Associationslexikon. It is
developed at Spr{\aa}kbanken at Gothenburg University
and intended for usage in language technology
research. 
We have developde tools for extracting GF lexicons
from SALDO, described in \ref{sec:extractsaldo}
%which have resulted in a dictionary containing
%over 98000 entries including information about verbs that
%are reflexive or needs a particle.
%The importing method should be fast and reliable
%enough to allow us to always have a fresh version of the dictionary
%in GF.

\section{Related work}
%Copied form thesis!!!! 
% Add: how are these related to this work?
% Add :detaild comparison of outputs of various parsers of Sw.
Many years of research have lead to many interesting language
technology tools for Swedish.
An example is the well-known data-driven Malt parser \cite{malt},
trained on Talbanken. 
There are also a number of grammar-based parsers, although none is freely available.
The cascaded finite state parser CassSwe \cite{casswe} and
The Swedish Constraint Grammar \cite{birn}
give syntactic analyses. 
Swedish FDG (Voultanien,2001) uses the Functional Dependency Grammar
\cite{fdg},
an extension of the Constraint Grammar
formalism, and produces a dependency structure focusing on finding the nominal
arguments. \\

The LinGO Grammar Matrix \cite{matrix}, is a starter-kit for building
Head-Driven Phrase Structure Grammars \cite{hpsg} (HPSG) providing
compatibility with tools for parsing, evaluation, semantic representations etc.
Translation is supported by using Minimal Recursion
Semantics \cite{mrs} as an interlingua. 
There is a collection of grammars implemented in this framework, giving broad-coverage
descriptions of 
English, Japanese and German. 
The Scandinavian Grammar Matrix \cite{scandmatrix} covers common parts of
Scandinavian, while Norsource (Hellan, 2003) describes Norwegian. A Swedish version
was based upon this by Ahrenberg, covering the morphology and some
differences between Swedish and Norwegian. Further, there is the BiTSE 
grammar \cite{stymne}, also implemented using the Lingo Matrix,
which focuses on describing and translating verb frames.\\

@@

\section{Extracting a large lexicon}
\label{sec:extractsaldo}
The lexicon provided with the GF resources is far too small for open-domain
parsing.
This section describes the process of importing SALDO, which is 
compatible with GF, and easily translated to GF format.
As SALDO is continuously updated, the importing process has been designed to be fast
and stable enough to be redone at any time.

\subsubsection{Implementation}
The basic algorithm for importing SALDO was implemented by Angelov (2008)
and  produces code for a GF lexicon.
For each word in SALDO, it decides which forms should be used as input
to the GF smart paradigms. The smart paradigm is a function which given one
form of a word, can infer which paradigm it most likely belongs to.
For verbs, this will in most cases mean giving
the present tense form, see figure \ref{fig:saldoknyt}. \\

@@

All assumed paradigms are printed to a temporary lexicon, 
which will produce an inflection table for every entry when compiled.
The tables are compared to the information given
in SALDO and if the tables are equal the code for the word is saved. If the table
is erroneous, another try is made
by giving more forms to the smart paradigm.
For example \ref{fig:saldoknyt}, the smart paradigm will fail to calculate the
correct inflection table. In the next try both the present and the past tense
are given:\\

\begin{figure}[h]
\begin{center}
\verb-mkV "knyter" "knöt" ;-
\caption{Second output for the verb `knyta'}
\label{fig:saldoknyt2}
\end{center}
\end{figure}

The program is run iteratively until the GF table matches the one given in SALDO,
or until there are no more ways of using the smart paradigm. The verb 'knyta'
will need three forms:\\

\begin{figure}[h]
\begin{center}
\verb-mkV "knyter" "knöt" "knutit"-\\
\caption{Final output for the verb `knyta'}
\label{fig:saldoknyt3}
\end{center}
\end{figure}

\subsubsection{Results}
\label{sec:saldoRes}
The resulting dictionary contains more than 100 000 entries, approximately 80 \% 
of the total size of SALDO.
There are a number of reasons why some words were not imported,
the most obvious one is that we do not want all categories from
SALDO in the GF lexicon. Prepositions, 
numerals, 
personal pronouns etc.
are assumed to be present in the resource grammars and should not be added again.
SALDO contains many pronouns which
are not analysed the same way in GF. %TODO see where??
Before adding them to our lexicon, we need to do more analysing to find their
correct GF-category. 
%Some experiments on finding the category
%have been done using Talbanken, see section \ref{sec:gf.quant}.
Categories involving multiple words 
are usually handled as idioms and should be given in a separate lexicon. In
total six types of words were considered for the extraction: \\

\begin{figure}[h]
\begin{tabular}{|l|ll|}
\hline
& SALDO$/$GF & Example \\
\hline
 Adverb & \textbf{ab}$/$\textbf{Adv} & ofta (\emph{often})\\
 Adjective&\textbf{av}$/$\textbf{A} & gul (\emph{yellow})\\
 Noun & \textbf{nn}$/$\textbf{N} & hus (\emph{house})\\
 Verb & \textbf{vb}$/$\textbf{V} & springa (\emph{run})\\
 Reflexive verbs &\textbf{vbm}$/$\textbf{V} & raka sig (\emph{shave})\\
 Particles verbs &\textbf{vbm}$/$\textbf{V}  &  piggna till (\emph{perk up})\\
\hline
\end{tabular}
\caption{}
\end{figure}

Most but not all words of these categories have been imported.
One reason why the importing phase would fail 
is that SALDO, unlike GF, only contains the actually used word forms.
For technical reasons, the smart paradigm might need forms never used.
Consider for example the plural tantum 
noun \emph{`glas{\"o}gon'} (\emph{`glasses'}).
The smart paradigm requires a singular form, and since the program could not
find this in SALDO, there was no way of adding the lemma to the lexicon. 
When the program failed to import a noun, this was often the explanation.
Words of this type may be added manually, for \emph{`glas{\"o}gon'} we could use
the ostensibly correct singular form `glas{\"o}ga', although this
has another meaning (`glass-eye').
The same problem occurred for the irregular s-verbs,
(\emph{`synas'} \emph{(`show')} or \emph{umg{\aa}s} \emph{(`socialize')})
which made up 61.5 \% of the failing verbs of type \verb_vb_.\\
In a few cases the smart paradigms could not generate the correct declination.\\

When testing the coverage of Talbanken,
we found that there are around 2500 word forms still missing, excluding the ones
tagged as names and numbers. This number may seem very high, but 
4/5 of the word forms are compounds and when performing the intended parsing,
an additional analysis identifying compounds is preformed before
looking-up the words in the lexicon. We
should also take into consideration that 
we cannot automatically find out how many actually stem from the same word, or
how many abbreviations that are present. Talbanken also contains a small number of 
spelling errors, which probably are enumerated among our missing words. The majority
of the missing words are only used once.\\

@@

A list of words that were given different labels in GF than in Talbanken has been
composed, consisting of about 1600 entries. Many of those are
acceptable and reflects the difference 
made in the analyses, while
others are examples of words that are still missing from
the lexicon.

%Since SALDO does not give any valency information, neither does the imported lexicon.
%The lexicon does not contain any information.
\noindent Valency information, which is crucial for GF, is not given in SALDO and
hence not in the imported lexicon. Instead we extract this from
Lexin\footnote{http://spraakbanken.gu.se/lexin/}.

\subsection{A tool for lexical acquisition}
As we extract our main lexicon from SALDO, we have also created a tool for
semi-automatic acquisition to complement the lexicon.
%We have tested it on verbs, with good results.
Like the SALDO importer it makes use of
the smart paradigm given in the Swedish resource grammar.
Unlike the SALDO importer, this tool does not require any particular
word form, but operates on any verb form and 
%By combining this method with the information from the tags in Talbanken,
%the tool interactively generates GF lexical entries. 
%Given a list of words, it 
iteratively tries to figure out how to conjugate each of them. If several forms
of a word are 
given, the program will try to identify the one that carries the most linguistic
information, put this in a form recognized by the smart paradigm and ask GF to output
a table with the resulting inflection. 
If the table contains all other conjugations from the input list,
the program will ask the  user to
validate the claimed paradigm. 
This step is needed since the input may not provide information enough
to automatically do the validation. The user may now either
allow the word to be added to the lexicon, remove it or request another guess.
The user hence only needs to decide if each paradigm is correct or not, and
the program will do the rest.

The tool has been tested on verbs. Although using simple techniques it 
manages to assign the correct paradigm to 70-75 \% of the given lemmas.
A smaller test has shown that out of the accepted lemmas, the correct guess is
made directly in 75\% of the cases, whereas the user has to reject one or more
guesses for 25\%. 
%Hence the program gets a correctness rate of 56 \% if it is
%run without help from the user. This kind of use is however not intended,

@@

\section{Translating The Treebank Talbanken}
\label{sec:mapping}
The information from the tags in Talbanken can be used for many purposes.
We have developed an automatic transformation of Talbanken trees 
to trees in GF format. The translation makes use of the POS tags as well as
the syntactic information. 
\begin{figure}[h!]
\begin{center}
\hspace{-30mm}
\subfloat{\label{pic:tbtree}\includegraphics[width=60mm]{Talbankentree.pdf}}
\hspace{-10mm}
\subfloat{\label{pic:gftree}\includegraphics[width=90mm]{gfTree.pdf}}
%\subfloat{\label{pic:gfptree}\includegraphics[width=50mm]{gfPTree.pdf}}
\caption{The Talbanken tree and the GF tree for the sentence ``Katten p{\aa} bilen blir st{\"o}rre".}
\label{fig:translationtrees}
\end{center}
\end{figure}

Figure \ref{fig:translationtrees} shows an example of a visualized Talbanken05 tree
of the sentence ``Katten p{\aa} bilen blir st{\"o}rre" 
(\emph{``The cat on the car gets bigger"}).
The mapping is based on 
a translation of the English Penn Treebank \cite{gfpenn}.
By modifying this program, we now have a translation that works for
the Tiger XML-format of
Talbanken. Adaption was required for the differences in annotation as well as 
for the syntactic differences between Swedish and English.

The translation %connects GF with another annotations and 
gives us means to evaluate our parser. By both parsing a Talbanken sentence and
transforming its annotated tree, we can easily inspect if the results are
equal.
Additionally, the mapping shows which grammatical constructions that are still missing
from the GF grammar and shows how the GF analysis differs from the one made
in Talbanken.
If there are words missing from our dictionary, the rich
POS-tags may help us to automatically find the correct declination and add it to the
lexicon. Further, our parser will need probabilities of how often a function is
used. The GF treebank we achieve from the translation is a good source for this
information.\\

%The mapping gives information about which form a word is
%currently used in and this may be used by the lexical extraction
%tools.
%The translated trees enable us to extract probabilities for how often
%different functions are used, a feature that will enable disambiguation.
%Furthermore, the translation makes it easy to identify grammatical constructions
%missing from the GF grammar and shows how the GF analysis differs from the one made
%in Talbanken.
%Another important use of the mapping is evaluation of the parser, which can be
%accomplished by comparing the parse trees and the trees from the transformer.
%\subsection{Differences in the notation}
%Some structural differences between GF and Talbanken became obvious during the
%translation.
%In Talbanken the valency information is given implicitly by the complements
%of a word. If a verb is followed by an object, \verb-OO-, which contains an
%\verb-S-, we can conclude that this is a sentence-complement verb.
%In GF, the valency information is given for each entry in the lexicon.
%A sentence-complement verb has the type \verb-VS- and from this we know 
%that the function \verb-ComplVS- should be used to combine the verb and with a 
%sentence.
%\begin{wrapfigure}{l}{0.52\textwidth}
%\includegraphics[width=80mm]{metatree.pdf}
%\caption{Abstract tree with meta variable}
%\label{pic:gMetatree}
%\end{wrapfigure}
%
%\noindent A similar issue arises from the use of verb particles. A verb with a
%particle is treated as two independent units in Talbanken but in GF they are
%considered to be one constituent. % which means that 
%The particle has therefore no effect on the abstract tree, its presence is
%announced by the verb \\itself.
%
%
%We have so far concentrated on shorter sentences, without idioms and
%conjunction. 
%Since GF demands information 
%that is not given by the Talbanken tags -- such as the valency of verbs -- a
%full translation requires that all words used are already known to GF. 
%The evaluation of trees of this sort shows that the mapping
%can restore 85\% of the nodes. 
%If we lift the restrictions, we can restore 63\% of the tags.
\subsection{Results and evaluation}
%>> more! replace words, how much better results (both grammar and mapping) 
%>> more results
%>> motivation, why so bad??\\
The development was mostly example-driven, and at least one rule for the translation of
every Talbanken-tag has been implemented.
Shorter sentences have been prioritized in order to get a coverage of the most
fundamental constructions. 
Regression testing was used to ensure that a new rule would not violate the old ones
or allow trees to be translated erroneously.

The flat version of Talbanken05 was used for developing the mapping, but
the program works for both the flat and the deep Tiger-XML versions.
%The deeper contains six more tags, although they are not always useful for our       
%purpose. Figure \ref{fig:mappDeepFlat} shows the difference for a conjunction.
%% show piece of s5 here!
%\begin{figure}[h]
%\begin{parsetree}
%    ( .CNP. 
%        (.CJ.  (.NP. ~ `oskifta dödsbon'))
%        ( .++. (.++OC. `och'))
%        (.CJ.  (.VN. `familjestiftelser'))
%    )
%\end{parsetree}
%\begin{parsetree}
%    ( .CNP. 
%        (.CJ.  (.NP. ~ `oskifta dödsbon'))
%        ( .CC.
%            (.CNP. 
%                (.++.  (.++OC. `och' ))
%                (.C+.  (.VN. `familjestiftelser'))
%         ))
%    )
%\end{parsetree}
%\caption{An example of the difference in the two analyses}
%\label{fig:mappDeepFlat}
%\end{figure}
The focus on the mapping has been simple sentences, in order to cover the most
fundamental structures. 
%If the program is extended and adapted to cover complex
%conjunction, a deeper evaluation should be preformed to see which version that
%suits our needs the best.
%The deeper implementation also gives 
%some more information about verb phrases, by the tag \verb|VG| (\emph{verb group}),
%which groups an infinitival verb and its object together into a \verb|VP|.
%This information can however be extracted from the flat implementation, and the results
%get slightly better when using this version. \\

When evaluating the mapping, the results strongly depend on which restrictions we
put on the input. 
One of the reasons why a node cannot be translated, is the 
use of the tags show in figure \ref{fig:mapBadtag}.
The \textbf{PU} tag is used for graphic listings, and not for fluent text.
In our grammar there is naturally no corresponding function; 
%We are not interested in sentences made up by list, since 
the listings are meant for making the text look nice in
folders etc and outside the scope for the grammar itself. The tags \textbf{XX}
and \textbf{NAC} are often used since Talbanken makes a
difference between subject
and object noun phrases. %(which is not needed in GF (see section \ref{sec:swegf}))
The analysis of elliptical expression in (\ref{sent:krav})
\enumsentence{F\"or stora krav.\\
``Too high demands."}\label{sent:krav}
contains the tags \textbf{XX} and \textbf{NAC}, since it is not obvious
whether the noun phrase is used as subject or an object.
The tags shown in figure \ref{fig:mapBadtag} occur quite frequently in the treebank and are always translated
to metas, which lowers our result. \\
%How we cannot cover all cases. 
\begin{figure}[h]
\begin{tabular}{ll}
\textbf{NAC} & Not a constituent\\
\textbf{XP} & Other (non-coordinated) phrase\\
\textbf{DB} & Doubled function\\
\textbf{PU} & List item\\
\textbf{XX} & Unclassifiable part-of-speech\\
\end{tabular}
\caption{}\label{fig:mapBadtag}
\end{figure}
